Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-11-29 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revised abstract
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2905 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5b3b5277e9fa1e6f4167208a94c2646237fe0d8d

\Abstract{
Establishing a proper regression model for circular data such as animal movement
directions, periodic measurements of neuronal activity or wind directions has shown
to be very challenging. Commonly applied generalized linear and additive models 
can provide a useful toolbox, however, non-additive effects might not be detected 
and difficulties may arise in modeling smooth effects due to highly non-linear link 
functions and increasing irregularities when moving closer to the ``origin point'' 
(0 which is equal to $2\pi$ in the radiant case).
%% FIXME: clear explanation of difficulties
Therefore, we propose a new approach by applying distributional trees and 
forests to circular data. In particular, the parameters of the employed von Mises 
distribution are regressed on covariates by a tree-based model yielding a constant set 
of estimated distribution parameters for each resulting subgroup, regardless of any 
``circular shift''/the ``circular position''.
%As circular data appears in many different areas, the development of
%appropriate statistical models which can account for periodicity has become an
%important but challenging research field. For probabilistic modeling of
%circular data typically the von Mises distribution is employed. In order to
%regress its parameters on covariates, a circular regression tree is
%proposed as an alternative to more commonly-used generalized linear and additive 
%models. 
The resulting circular regression trees are easy to interpret, can detect non-additive effects, 
and select covariates and their interactions automatically. By combining an
ensemble of circular regression trees a more stable and regularized circular
regression forest can be built which allows for modeling smooth effects as
well.

To evaluate the proposed circular regression trees and forests, a case study on
probabilistic forecasting of wind direction for two airports in Austria is
presented. In addition to the tree-based models, alternative common approaches
are used as a benchmark in order to assess the predictive performances of
circular regression trees and forests. 
}
\Keywords{Distributional Trees, Random Forests; Circular Data; von Mises Distribution}

@@

due to the highly non-linear link function. Consequently, the representation of smooth transitions 
on the unit circle is not straightforward. In addition, the same rotation can be obtained in either 

@@

As a very intuitive and data-driven alternative, we propose a flexible
tree-based regression approach for modeling circular data by applying the von
Mises distribution within the methodology of distributional trees and
distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
%In order to avoid these difficulties in regression modeling of circular data 
%(and also the requirement of initial values) 
%we propose an alternative and very flexible 
%approach by applying the von Mises distribution within the methodology of distributional 
%trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
The resulting circular regression trees and forests avoid the discussed difficulties 
of circular GLMs by separating the data into more homogeneous subgroups and fitting a 
simple intercept-only model for each resulting subset.
Therefore, no covariates and, thus, no link function must be employed in the estimation
of the group-specific distribution parameters.
By incorporating the framework of distributional modeling we allow for detecting 
splits in all distribution parameters and provide a fully specified circular response distribution 
for each terminal node offering a wide range of inference.
In addition, the employed tree structure captures non-additive effects while forests allow for 
modeling smooth changes as well.
Furthermore, covariates and their possible interactions do not need to specified in advance as they are
selected automatically.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This novel approach to circular regression trees and forests complements the literature 
on tree-based circular modeling.
%In particular, 
\cite{Lund:2002} has already introduced the concept of circular regression trees where 
binary splits are made based on an angular distance measure employed to evaluate node 
homogeneity.
While this approach and also further extensions \citep{Larraondo+Inza+Lozano:2018} allow not 
only for linear but also for a circular covariate, the tree only considers effects in the mean 
direction for splitting, i.e., only the location parameter $\mu$ is regressed on covariates.
Therefore, the introduced distributional approach to circular regression trees extends the concept
of previous circular trees by considering splits in all distribution parameters and 
yielding a full probabilistic model. In addition, as a natural further development ensembles or forests 
of circular regression trees are presented in this study. 
% that can improve forecasts by regularizing and stabilizing the model.

@@

\mu = \mu_0 + g(\mathbf{\beta}^{\top} \mathbf{z})

@@

the additive predictor is transformed to an interval of length $2\pi$  by a link function $g$. 

@@

homogeneity, \citet{Fisher+Lee:1992} also provide a heteroscedastic version by combining two  

@@

The main goal of employing tree-based models to circular data is to provide a flexible and 
easy-to-apply modeling approach which accounts for a proper representation on the unit circle.
Based on ideas of \citet{Lund:2002}, we introduce probabilistic circular regression trees and forests
employing the von Mises distribution. The resulting tree-based models provide a very intuitive and data-driven 
alternative to commonly used circular regression models for circular data.

@@

\subsection{Circular forests}
\label{sec:circforest} 
A natural extension of circular regression trees are ensembles or forests
that can improve forecasts by regularizing and stabilizing the model. Random
forests introduced by \cite{Breiman:2001} average the predictions of an
ensemble of trees, each built on a subsample or bootstrap of the original data.
A generalization of this strategy is to obtain weighted predictions by adaptive
local likelihood estimation of the distributional parameters
\citep[Section~2.3. of][]{Schlosser+Hothorn+Stauffer:2019,
Hothorn+Zeileis:2017}. More specifically, for each possibly new
observation~$\bm{z}$ a set of averaged ``nearest neighbor'' weights
$w_i^{forest}(\bm{z})$ is obtained that is based on the number of trees in
which $\bm{z}$ is assigned to the same terminal node as each learning
observation $y_i, i \in \{1,\ldots,n\}$. Hence, for a forest of $T$ trees the
weights are calculated via
%%
\begin{equation}
  w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
  \frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
\end{equation}
%%
where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
segment of the $t$-th tree. Therefore, similar observations ending up more
often in the same terminal node have higher weights and in that way a stronger
influence in the estimation process.

>>>>

\Abstract{
Despite the occurrence of circular data in a wide range of scientific
fields, the existing methodology for probabilistically forecasting
circular data
is highly limited. Most of the existing methods are built on the
framework of
generalized linear and additive models, which can incorporate
difficulties in
both the optimization and the interpretation. As a very intuitive
alternative
to additive circular regression models, we extend previous ideas of circular
trees and introduce probabilistic circular regression trees and forests
employing the von Mises distribution. The resulting tree-based models
simplify
the estimation process by separating the data into more homogeneous
subgroups
and fitting a simple intercept-only model for each resulting subset. By
incorporating the framework of distributional modeling, splits in all
distribution parameters are considered and a fully specified distribution is
provided for each node.  The proposed circular regression trees are easy to
interpret, can detect non-additive effects, and select covariates and their
interactions automatically. By combining an ensemble of circular regression
trees a more stable and regularized circular regression forest can be built
which allows for modeling smooth effects as well.

To evaluate the proposed circular regression trees and forests, a case
study on
probabilistic forecasting of wind direction for two airports in Austria is
presented. In addition to the tree-based models, alternative common
approaches
are used as a benchmark in order to assess the predictive performances of
circular regression trees and forests.
%Establishing a proper regression model for circular data such as animal movement
%directions, periodic measurements of neuronal activity or wind directions,
%requires specific attention regarding its representation on the unit circle.
%Commonly applied generalized linear and additive models can provide a useful toolbox, 
%however, highly non-linear link functions can complicate interpretations while
%a non-concave log-likelihood function of the usually employed von Mises distribution 
%leads to difficulties in the incorporated optimization procedure.
%Moreover, an adequate model specification can be very challenging and non-additive 
%effects might not be detected.

%Therefore, an alternative approach to modeling circular data is proposed by
%employing the von Mises distribution within the methodology of distributional 
%trees and forests. Exploiting the tree-based separation of the data set into 
%subgroups an intercept-model is fitted within each subset such that no covariates 
%are included simplifying ....
%difficulties may arise in 
%modeling smooth effects due to highly non-linear link 
%functions and increasing irregularities when moving closer to the ``origin point'' 
%(0 which is equal to $2\pi$ in the radiant case).
%%% FIXME: clear explanation of difficulties
%Therefore, we propose a new approach by applying distributional trees and 
%forests to circular data. In particular, the parameters of the employed von Mises 
%distribution are regressed on covariates by a tree-based model yielding a constant set 
%of estimated distribution parameters for each resulting subgroup, regardless of any 
%``circular shift''/the ``circular position''.
%As circular data appears in many different areas, the development of
%appropriate statistical models which can account for periodicity has become an
%important but challenging research field. For probabilistic modeling of
%circular data typically the von Mises distribution is employed. In order to
%regress its parameters on covariates, a circular regression tree is
%proposed as an alternative to more commonly-used generalized linear and additive 
%models. 
%Moreover, the resulting circular regression trees are easy to interpret, 
%can detect non-additive effects, and select covariates and their interactions automatically. 
%By combining an ensemble of circular regression trees a more stable and regularized circular
%regression forest can be built which allows for modeling smooth effects as
%well.

%To evaluate the proposed circular regression trees and forests, a case study on
%probabilistic forecasting of wind direction for two airports in Austria is
%presented. In addition to the tree-based models, alternative common approaches
%are used as a benchmark in order to assess the predictive performances of
%circular regression trees and forests. 
}
\Keywords{Distributional Trees, Random Forests; Circular Data; von Mises Distribution}

@@

due to the highly non-linear link function and the representation of smooth transitions 
on the unit circle is not straightforward. For example, the same rotation can be obtained in either 

@@

As a very intuitive and data-driven alternative, we propose a flexible
tree-based regression approach for modeling circular data by applying the von
Mises distribution within the methodology of distributional trees and
distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
%In order to avoid these difficulties in regression modeling of circular data 
%(and also the requirement of initial values) 
%we propose an alternative and very flexible 
%approach by applying the von Mises distribution within the methodology of distributional 
%trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
The resulting circular regression trees and forests avoid the discussed difficulties 
of circular GLMs by separating the data into more homogeneous subgroups and fitting a 
simple intercept-only model for each resulting subset.
Therefore, no covariates and, thus, no link function must be employed in the estimation
of the group-specific distribution parameters which also simplifies the optimization process.
By incorporating the framework of distributional modeling, we allow for detecting 
splits in all distribution parameters and provide a fully specified circular response distribution 
for each terminal node offering a wide range of inference.
In addition, the employed tree structure captures non-additive effects while forests allow for 
modeling smooth changes as well.
Furthermore, covariates and their possible interactions do not need to be specified in advance as 
they are selected automatically.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This novel approach to circular regression trees and forests complements the literature 
on tree-based circular modeling.
%In particular, 
\cite{Lund:2002} has already introduced the concept of circular regression trees where 
binary splits are made based on an angular distance measure employed to evaluate node 
homogeneity.
While this approach and also further extensions \citep{Larraondo+Inza+Lozano:2018} allow not 
only for linear but also for a circular covariate, the tree only considers effects in the mean 
direction for splitting, i.e., only the location parameter $\mu$ is regressed on covariates.
Therefore, the introduced distributional approach to circular regression trees extends the concept
of previous circular trees by considering splits in all distribution parameters and 
yielding a full probabilistic model. In addition, as a natural further development, ensembles or forests 
of circular regression trees are presented in this study. 
% that can improve forecasts by regularizing and stabilizing the model.

@@

\mu = \mu_0 + g\left(\mathbf{\beta}^{\top} \mathbf{z}\right),

@@

additive predictor is transformed to an interval of length $2\pi$  by a link function $g$. 

@@

homoscedasticity, \citet{Fisher+Lee:1992} also developed a heteroscedastic version by combining two  

@@

The main goal of employing tree-based models to circular data is to provide a flexible and 
easy-to-apply modeling approach which accounts for a proper representation on the unit circle.
Based on ideas of \citet{Lund:2002}, we introduce probabilistic circular regression trees and forests
employing the von Mises distribution. The resulting tree-based models provide a very intuitive and data-driven 
alternative to commonly used regression models for circular data.

@@

\subsection{Circular regression forests}
\label{sec:circforest} 
A natural extension of circular regression trees are ensembles or forests
that can improve forecasts by regularizing and stabilizing the model. Random
forests introduced by \cite{Breiman:2001} average the predictions of an
ensemble of trees, each built on a subsample or bootstrap of the original data.
A generalization of this strategy is to obtain weighted predictions by adaptive
local likelihood estimation of the distributional parameters
\citep[Section~2.3. of][]{Schlosser+Hothorn+Stauffer:2019,
Hothorn+Zeileis:2017}. More specifically, for each possibly new
observation~$\bm{z}$ a set of averaged ``nearest neighbor'' weights
$w_i^{forest}(\bm{z})$ is obtained that is based on the number of trees in
which $\bm{z}$ is assigned to the same terminal node as each learning
observation $y_i, i \in \{1,\ldots,n\}$. Hence, for a forest of $T$ trees the
weights are calculated via
%%
\begin{equation}
  w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
  \frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
\end{equation}
%%
where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
segment of the $t$-th tree. Therefore, similar observations ending up more
often in the same terminal node have higher weights and in that way a stronger
influence in the estimation process.
